Restructuring Data for Exascale Scientific Productivity

### 2016-03-11 papers/2016-HotStorage/paper.tex
finished paper.

%for single author (just remove % characters)
%\author{
%{\rm Your N.\ Here}\\
%Your Institution
%\and
%{\rm Second Name}\\
%Second Institution
% copy the following lines to add more authors
% \and
% {\rm Name}\\
%Name Institution
%} % end author
%\numberofauthors{1}
\author{
Jay Lofstead\footnotemark[1]~~,
Hasan Abbasi\footnotemark[3]~~,
Mark Ainsworth\footnotemark[6]~\footnotemark[3]~~,
Jong Choi\footnotemark[3]~~,
Matthew Curry\footnotemark[1]~~,\\
Scott Klasky\footnotemark[3]~\footnotemark[4]~\footnotemark[7]~~,
Tahsin Kurc\footnotemark[9]~~,
Qing Liu\footnotemark[3]~~,
Jeremy Logan\footnotemark[4]~~,
Carlos Maltzahn\footnotemark[2]~~,
Manish Parashar\footnotemark[5]~~,\\
Norbert Podhorzski\footnotemark[3]~~,
Feiyi Wang\footnotemark[3]~~,
Matthew Wolf\footnotemark[7]~~,
C. S. Chang\footnotemark[8]~~,
Michael Churchill\footnotemark[8]~~,
Stephane Ethier\footnotemark[8]~~
\\
\footnotemark[1]~~Sandia National Laboratories,
\footnotemark[2]~~University of California Santa Cruz,\\
\footnotemark[3]~~Oak Ridge National Laboratory,
\footnotemark[4]~~U. Tenn. Knoxville,
\footnotemark[5]~~Rutgers University,\\
\footnotemark[6]~~Brown University,
\footnotemark[7]~~Georgia Tech,
\footnotemark[8]~~PPPL,
\footnotemark[9]~~Stony Brook
}

%\author{
%\alignauthor {Jay Lofstead\footnotemark[1], Carlos Maltzahn\footnotemark[2], Scott Klasky\footnotemark[3], Hasan Abbasi\footnotemark[3], Qing Liu\footnotemark[3]\\
%Matthew Curry\footnotemark[1], Mark Ainsworth\footnotemark[5], Manish Parashar\footnotemark[4]}\\
%\footnotemark[1]~~Sandia National Laboratories
%\footnotemark[2]~~University of California\\
%\footnotemark[3]~~Oak Ridge National Laboratory
%\footnotemark[4]~~Rutgers University
%\footnotemark[5]~~Brown University\\
%}

@@

\subsection*{Abstract}
The longest existing parallel I/O APIs, HDF5, PnetCDF, and NetCDF, are focused
on treating the entire output from a paralell application as a single unit.
These libraries work to gather the data from all processes and arrange it such
that it is organized much like as if it were written from a single process.
This approach, in particular the two-phase collective I/O with data sieving
portion, has proven difficult to scale for 3-D domain decompositions at scale.
ADIOS shifted the focus from treating the entire application output as a single
entity to treating the output from each writer as a self-contained entity. This
eliminated the need for the problematic two-phase collective I/O and has
demonstrated good scalability well into the petascale era. Another shift is
required to address the needs of exascale systems. New system I/O bandwidth is
not keeping pace with the compute acceleration leading to accute bottlenecks.
While ``burst buffer'' technology will help address the performance gap, it
does not address the problem completely. This paper describes a new approach
to not just an I/O library, but also the underlying storage infrastructure to
address the needs of exascacle applications cognizant of projected exascale
platform characteristics and in the context of current industry trends.

@@

High performance computing simulations have driven tightly coupled platform
scaling for decades. The output from these simulations are analyzed using
regular tools and techniques optimized for this environment.  The first
generation of standardized APIs for scientific computing IO focused on serial
output from a single process to a distributed or local file system.  These
systems still exist in the form of serial
HDF5~\cite{chilan:2006:hdf5-performance} and the end-of-life NetCDF
3.x~\cite{netcdf3}. The key feature of these APIs was the logical data model
that eliminated any parallel data decomoposition artifacts from the actual data
storage. This offered consistent performance and regular data access patterns.

@@

This paper presents an overview of the efforts to build such a fourth
generation storage and IO system. A discussion of the end-user API, storage
challenges, quality of service features, and metadata challenges are all
discussed briefly.

@@

\noindent\textbf{Refactoring} may be a uniform operation across an
entire variable. For example, splitting a double into three parts--two byes,
four bytes, and two bytes, yields a two-byte value that is the majority of the
data precision. The next four bytes would be most of the rest of the data
precision. The last two bytes are the remainder of the precision. For quick
overviews, just reading the two-byte portion is sufficient to get a rough idea
of data contents. If particular data looks interesting, more and more data can
be read to improve the precision. This can easily be applied uniformly by each
process without coordinating. From a storage hierarchy perspective, the primary
value portion could be written to high performance, low capacity storage while
the rest is written to slower storage. That would reduce the data quantity
written to NVM, for example, by 75\% for a 2/6 split enabling far more output
to be written to this fast tier with the rest spooled asynchronously.

A more complex, but interesting approach is to offer an {\em Auditor} that
performs a similar, but lower precision computation that greatly reduces
computation intensity with only a minor loss of precision over short spatial
and temporal domains. As the simulation progresses, the auditor drift would
need to be reset to bring it back into alignment. The much smaller, reduced
precision data could be used instead of the full precision in many
circumstances.  A more dynamic lossy approach would be to apply a wavelet
compression~\cite{klappenecker:1995:wavelet} approach that generates different
compressed data sizes at a fixed error rate depending on the data entropy.
These are still uniform operators, but the resulting data can be of radically
different sizes. For this approach, some or all of the lower precision data
can be written to the fast tier while full precision can be spooled
asynchrnously to the slower tier(s).

A middle ground approach between these two refactoring approaches is to use a
non-uniform operator across a variable. Instead of, for example, splitting on
byte boundaries to make multiple values, a function that identifies data
features through a local analytics function could identify ``interesting'' data
that should be maintained in full precision while other areas can be more
highly compressed. Such an operator would decompose the data from a process
further than a single process or node depending on the data features. This
approach may balance precision better than wavelet compression, but at a higher
analytics cost. We would put data areas with interesting features on the fast
tier while the rest could be spooled to the slow tier asynchronously.

We plan to investigate all three approaches as different applications will have
different techniques that work best. Our initial tests have demonstrated the
potential for the auditor approach in an isolated environment. We have also
built a prototype ADIOS transport method for the byte-split apporach. Any
performant storage system should offer the flexibility to incorporate
application specific data refactoring techniques and sufficient metadata
support for an arbitrary client to access the data successfully.

\noindent\textbf{Quality of Service} deadlines are important for driving how
much time is spent performing IO. At the same time, data quality must be
maintained at a sufficient level for the resulting science to be possible.
The system must offer a way for an end-user to negotiate with the system for
trading off between time spent in IO and the amount of data written or read.
Part of that tradeoff will be increased or reduced data quality, depending on
the system. For example, when data is written, a few different forms may be
output given sufficient space and time. One may be every fourth element
across a regular mesh reduces the data size to $1/64^{th}$ the full data set.
Further dividing that using byte-splitting can reduce it by another 75\%. For
quick overviews to search data sets, this reduced form can be sufficient
driving which data sets should be investigated in more detail. The key driving
factor is time. The applicaiton scientist can read 0.4\% of the full data
volume with a similar time reduction. SIRIUS will offer a negotiation process
to read any data.

@@

First, a user will request a particular variable and timesetp. SIRIUS will
return a table similar to Table~\ref{tab:results}. For a given variable ``A'',
the end-user can then select the data quality and approximate reading time for
the operation. In this case, there are 5 options for obtaining the whole data
set (1/2/3/4, 5/6, 7/8, 9, 10). The first four represent a spatial and byte
split, 5/6 represent just strided writing, 7/8 represent just a byte-split, 9
represents a wavelet compressed version, and 10 is the full precision data.
The user would specify to read a particular data instance that would proceed
normally.  Quality of Service soft guarantees specified as part of the request
will result in SIRIUS allocating soft bandwidth reservations to meet the
request. While some of the error rates seem to render the data unusable,
consider a 3000x3000x3000 simulation domain. That would yield a 27 gigapixel
image should all of the data be used. Visualization will already radically
reduce the data sizes to make usable visualizations that work on a standard
display. A 99\% reduction would yield a 270 megapixel image--still quite large.

@@

SIRIUS offers a novel feature related to managing IO time as a first class
characteristic. Rather than being at the mercy of the storage array
performance, an end-user can describe the local data to the storage system and
receive an estimate on how long that data will take to write. If that is
acceptable within a margin of error, the API can write. If it is unacceptable,
the end-user can make decisions about skipping the output or maybe spending a
little time with a strong data refactoring to reduce the data size.

On reading, a user may specify that they are willing to wait for an approximate
time for an operation. This will negotiate with the storage system to determine
the data quality returned. If the requested data quality cannot be retrieved
within a given time, the storage array may be able to apply an additional
data reduction operator to decrease the amount of time spent in IO.
Alternatively, if multiple different refactored data versions are stored,
alternative versions can be requested instead.

@@

The metadata system has a daunting challenge. Traditional metadata systems for
both distributed and parallel file systems only have to deal with a single name
space and storage hierarchy. While hierarichal storage management~\cite{blaze:1992:hsm}
systems offer a way to show a single name space, the reality is that the tier
closest to the user is a cache for the other tiers with data migrating as
necessary to support user interaction. SIRIUS is working in a fundamentally
different way. Each tier is treated as a first class citizen with direct
client access possible. Where data is actually retrieved from is where it is
actually stored rather than forcing data migration and caching. Systems like
Sirocco~\cite{curry:2015:sirocco} offer a way to support this model directly and are a
fundamental part of our system.

@@

We use the XGC1~\cite{ku:2008:xgc1} fusion simulation as our application. We
run on 5 nodes with 160 processes total. XGC1 is configured to use 819,200,000
particles evenly spread across the processes.

@@

Overall, the SIRIUS storage system is ushering in a new generation of both IO
APIs and storage system managed together. It incorporates additional metadata
to help accelerate data access and ultimately time to insight. Our initial
efforts show promise that such a system is both possible and can offer
attractive performance to better use the supercomputing platform.

>>>>

\author{
Jay Lofstead\footnotemark[1]~~,
Hasan Abbasi\footnotemark[3]~~,
Mark Ainsworth\footnotemark[6]~\footnotemark[3]~~,
Jong Choi\footnotemark[3]~~,
Matthew Curry\footnotemark[1]~~,\\
Scott Klasky\footnotemark[3]~\footnotemark[4]~\footnotemark[7]~~,
Tahsin Kurc\footnotemark[9]~~,
Qing Liu\footnotemark[3]~~,
Jeremy Logan\footnotemark[4]~~,
Carlos Maltzahn\footnotemark[2]~~,
Manish Parashar\footnotemark[5]~~,\\
Norbert Podhorzski\footnotemark[3]~~,
Feiyi Wang\footnotemark[3]~~,
Matthew Wolf\footnotemark[7]~~,
C. S. Chang\footnotemark[8]~~,
Michael Churchill\footnotemark[8]~~,
Stephane Ethier\footnotemark[8]~~
\\
\footnotemark[1]~~Sandia National Laboratories,
\footnotemark[2]~~University of California Santa Cruz,\\
\footnotemark[3]~~Oak Ridge National Laboratory,
\footnotemark[4]~~U. Tenn. Knoxville,
\footnotemark[5]~~Rutgers University,\\
\footnotemark[6]~~Brown University,
\footnotemark[7]~~Georgia Tech,
\footnotemark[8]~~PPPL,
\footnotemark[9]~~Stony Brook
}

@@

\subsection*{Abstract}
The longest existing parallel I/O APIs, HDF5, PnetCDF, and NetCDF, are focused
on treating the entire output from a parallel application as a single unit.
These libraries work to gather the data from all processes and arrange it such
that it is organized much like as if it were written from a single process.
This approach, in particular the two-phase collective I/O with data sieving
portion, has proven difficult to scale for 3-D domain decompositions at scale.
ADIOS shifted the focus from treating the entire application output as a single
entity to treating the output from each writer as a self-contained entity. This
eliminated the need for the problematic two-phase collective I/O and has
demonstrated good scalability well into the petascale era. Another shift is
required to address the needs of exascale systems. New supercomputing system
I/O bandwidth is not keeping pace with the compute acceleration leading to
accute bottlenecks.  While ``burst buffer'' technology will help address the
performance gap, it does not address the problem completely. This paper
describes a new approach to not just an I/O library, but also the underlying
storage infrastructure to address the needs of exascacle applications cognizant
of projected exascale platform characteristics and in the context of current
industry trends.

@@

High performance computing simulations have driven tightly coupled platform
scaling for decades. The output from these simulations are analyzed using
regular tools and techniques optimized for this environment.  The first
generation of standardized APIs for scientific computing IO focused on serial
output from a single process to a distributed or local file system.  These
systems still exist in the form of serial
HDF5~\cite{chilan:2006:hdf5-performance} and the end-of-life NetCDF
3.x~\cite{netcdf3}. The key feature of these APIs is the logical data model
that eliminates any parallel data decomoposition artifacts from the stored
data.  This offered consistent performance and regular data access patterns.

@@

This paper presents an overview of the efforts to build such a fourth
generation storage and IO system. A discussion of the platform features,
storage challenges, quality of service features, and metadata challenges are
all discussed briefly.

@@

\subsection{Refactoring}
Refactoring may be a uniform operation across an entire variable. For example,
splitting a double into three parts--two byes, four bytes, and two bytes,
yields a two-byte value that is the majority of the data precision. The next
four bytes would be most of the rest of the data precision. The last two bytes
are the remainder of the precision. For quick overviews, just reading the
two-byte portion is sufficient to get a rough idea of data contents. If
particular data looks interesting, more and more data can be read to improve
the precision. This can easily be applied uniformly by each process without
coordinating. From a storage hierarchy perspective, the primary value portion
could be written to high performance, low capacity storage while the rest is
written to slower storage. That would reduce the data quantity written to NVM,
for example, by 75\% for a 2/6 split enabling far more output to be written to
this fast tier with the rest spooled asynchronously.

A similar approach that maintains full data quality is to use a data stride to
store only a data subset. For example, by selecting every fourth element in
three dimensions yields a $\frac{1}{64}$ sized data set to store in fast
storage--a mere 1.5\% of the original size. The rest can be spooled
asynchronously. Combining byte-splits and striding can reduce the data sizes
even more. Doing a 3/5 split and an every fourth element yields 0.9\% of the
original size.

A more complex, but interesting approach is to offer an {\em Auditor} that
performs a similar, but lower precision computation that greatly reduces
computation intensity with only a minor loss of precision over short spatial
and temporal domains. As the simulation progresses, the auditor drift would
need to be reset to bring it back into alignment. The much smaller, reduced
precision data could be used instead of the full precision in many
circumstances.  For an example auditor, consider a previous generation physics
model. While it is not as accurate as the state-of-the-art-model, it is very
close, particularly over short temporal and spatial ranges. Other types of
auditors are possible and are under investigation. The key idea with the
auditor is that the reduced quality result can be calculated much more quickly
than the full precision and will have much smaller data sizes. As a side
benefit, the auditor can both detect soft errors in the caluclation by the
drift from the high precision physics diverging too quickly. It can also be
used to create a statistically representative data set for the high precision
physics model. Calculating the auditor value and writing its output for
checkpoint restart or initial analysis exploration can save time and leverage
otherwise idle excess compute resources.

A more dynamic lossy approach would be to apply a wavelet
compression~\cite{klappenecker:1995:wavelet} approach that generates different
compressed data sizes at a fixed error rate depending on the data entropy.
These are still uniform operators, but the resulting data can be of radically
different sizes. For this approach, some or all of the lower precision data
can be written to the fast tier while full precision can be spooled
asynchrnously to the slower tier(s).

A middle ground approach between these refactoring approaches is to use a
non-uniform operator across a variable. Instead of, for example, splitting on
byte boundaries to make multiple values, a function that identifies data
features through a local analytics function could identify ``interesting'' data
that should be maintained in full precision while other areas can be more
highly compressed. Such an operator would decompose the data from a process
further than a single process or node depending on the data features. This
approach may balance precision better than wavelet compression, but at a higher
analytics cost. SIRIUS would put data areas with interesting features on the
fast tier while the rest could be spooled to the slow tier asynchronously.

We plan to investigate all of these approaches as different applications will
have different techniques that work best. Our initial tests have demonstrated
the potential for the auditor approach in an isolated environment. We have also
built a prototype ADIOS transport method for the byte-split apporach. Any
performant storage system should offer the flexibility to incorporate
application specific data refactoring techniques and sufficient metadata
support for an arbitrary client to access the data successfully.

\subsection{Quality of Service}
Quality of service deadlines are important for driving how much time is spent
performing IO. At the same time, data quality must be maintained at a
sufficient level for the resulting science to be possible.  The system must
offer a way for an end-user to negotiate with the system for trading off
between time spent in IO and the amount of data written or read.  Part of that
tradeoff will be increased or reduced data quality, depending on the system.
For example, when data is written, a few different forms may be output given
sufficient space and time. One may be every fourth element across a regular
mesh reduces the data size to $1/64^{th}$ the full data set.  Further dividing
that using byte-splitting can reduce it by another 75\%. For quick overviews to
search data sets, this reduced form can be sufficient driving which data sets
should be investigated in more detail. The key driving factor is time. The
applicaiton scientist can read 0.4\% of the full data volume with a similar
time reduction. SIRIUS will offer a negotiation process to read any data.

@@

To read these various data organizations, read operations will shift to a
two-step process. First, an list of options will be requested that includes
both and error estiates. A second request will retrieve a particular data
portion from the provided list. Consider Table~\ref{tab:results}.  First, a
user will request a particular variable and timesetp. SIRIUS will return a
table similar to Table~\ref{tab:results}. For a given variable ``A'',
the end-user can then select the data quality and approximate reading time for
the operation. In this case, there are 5 options for obtaining the whole data
set (1/2/3/4, 5/6, 7/8, 9, 10). The first four entries represent a spatial and
byte split, 5/6 represent just strided writing, 7/8 represent just a
byte-split, 9 represents a wavelet compressed version, and 10 is the full
precision data.  The user would specify to read a particular data instance that
would proceed normally.  Quality of Service soft guarantees specified as part
of the request will result in SIRIUS allocating soft bandwidth reservations to
meet the request. While some of the error rates seem to render the data
unusable, consider a 3000x3000x3000 simulation domain. That would yield a 27
gigapixel image should all of the data be used. Visualization will already
radically reduce the data sizes to make usable visualizations that work on a
standard display. A 99\% reduction would yield a 270 megapixel image--still
quite large.

@@

SIRIUS offers a novel feature related to managing IO time as a first class
characteristic. Rather than being at the mercy of the storage array
performance, an end-user can describe the local data to the storage system and
receive an estimate on how long that data will take to write. If that is
acceptable within a margin of error, the API can write. If it is unacceptable,
the end-user can make decisions about skipping the output or maybe spending a
little time with a strong data refactoring to reduce the data size.  On
reading, a user will select from a list of available options based on the time
and data quality requirements.

Key to making QoS work is to maintain performance metadata to inform the IO
time estimates and error bounds returned to the user. Without hard
reservations, no system can offer strict performance guarantees. Instead, by
monitoring system behavior, SIRIUS will be able to offer reasonable time
estimates given data placement and the requesting source(s).

@@

The metadata system has a daunting challenge. Traditional metadata systems for
both distributed and parallel file systems only have to deal with a single name
space and storage tier. While hierarichal storage
management~\cite{blaze:1992:hsm} systems offer a way to show a single name
space, the reality is that the tier closest to the user is a cache for the
other tiers with data migrating as necessary to support user interaction.
SIRIUS is working in a fundamentally different way. Each tier is treated as a
first class citizen with direct client access possible. Where data is actually
retrieved from is where it is actually stored rather than forcing data
migration and caching. Systems like Sirocco~\cite{curry:2015:sirocco} offer a
way to support this model directly and are a fundamental part of our system.

@@

We use the XGC1~\cite{ku:2008:xgc1} fusion simulation as our application. We
run on 5 nodes with 160 processes total. XGC1 is configured to use 819,200,000
particles evenly spread across the processes. Performance numbers are
presented in Table~\ref{tab:performance}.

@@

Overall, the SIRIUS storage system is ushering in a new generation of both IO
APIs and storage system managed together. It incorporates additional metadata
to help accelerate data access and ultimately time to insight. Our initial
efforts show promise that such a system is both possible and can offer
attractive performance to better use the supercomputing platform. By treating
all storage tiers as first-class resources accessed independently, SIRIUS can
maximize available IO bandwidth. By refactoring data to extract out important
parts into separate data set slices, data writing and reading time can be
greatly improved while still maintaining full data fidelity when necessary.
The quality of service monitoring will inform the performance estimates giving
a more predictably responsive system accelerating time to insight.

### 2016-03-11 papers/2016-Supercomputing/paper.tex
finished paper.

\numberofauthors{1}
\author{
\alignauthor {
Jay Lofstead\footnotemark[1]~~,
Hasan Abbasi\footnotemark[3]~~,
Mark Ainsworth\footnotemark[6]~\footnotemark[3]~~,
Jong Choi\footnotemark[3]~~,
Matthew Curry\footnotemark[1]~~,\\
Scott Klasky\footnotemark[3]~\footnotemark[4]~\footnotemark[7]~~,
Tahsin Kurc\footnotemark[9]~~,
Qing Liu\footnotemark[3]~~,
Carlos Maltzahn\footnotemark[2]~~,
Manish Parashar\footnotemark[5]~~,\\
Norbert Podhorzski\footnotemark[3]~~,
Feiyi Wang\footnotemark[3]~~,
Matthew Wolf\footnotemark[7]~~,
C. S. Chang\footnotemark[8]~~,
Michael Churchill\footnotemark[8]~~,
Stephane Ethier\footnotemark[8]~~
}\\
\footnotemark[1]~~Sandia National Laboratories,
\footnotemark[2]~~University of California Santa Cruz,\\
\footnotemark[3]~~Oak Ridge National Laboratory,
\footnotemark[4]~~U. Tenn. Knoxville,
\footnotemark[5]~~Rutgers University,\\
\footnotemark[6]~~Brown University,
\footnotemark[7]~~Georgia Tech,
\footnotemark[8]~~PPPL,
\footnotemark[9]~~Stony Brook
}
\maketitle

>>>>

\numberofauthors{1}
\author{
\alignauthor {
Jay Lofstead\footnotemark[1]~~,
Hasan Abbasi\footnotemark[3]~~,
Mark Ainsworth\footnotemark[6]~\footnotemark[3]~~,
Jong Choi\footnotemark[3]~~,
Matthew Curry\footnotemark[1]~~,\\
Scott Klasky\footnotemark[3]~\footnotemark[4]~\footnotemark[7]~~,
Tahsin Kurc\footnotemark[9]~~,
Qing Liu\footnotemark[3]~~,
Jeremy Logan\footnotemark[4]~~,
Carlos Maltzahn\footnotemark[2]~~,
Manish Parashar\footnotemark[5]~~,\\
Norbert Podhorzski\footnotemark[3]~~,
Feiyi Wang\footnotemark[3]~~,
Matthew Wolf\footnotemark[7]~~,
C. S. Chang\footnotemark[8]~~,
Michael Churchill\footnotemark[8]~~,
Stephane Ethier\footnotemark[8]~~
}\\
\footnotemark[1]~~Sandia National Laboratories,
\footnotemark[2]~~University of California Santa Cruz,\\
\footnotemark[3]~~Oak Ridge National Laboratory,
\footnotemark[4]~~U. Tenn. Knoxville,
\footnotemark[5]~~Rutgers University,\\
\footnotemark[6]~~Brown University,
\footnotemark[7]~~Georgia Tech,
\footnotemark[8]~~PPPL,
\footnotemark[9]~~Stony Brook
}
\maketitle
